\chapter{Theory}

In this chapter, I am going to describe the theories behind our experiment. It is divided into three parts. The first section explains the theories of cold Bose gas and the Bose-Einstein condensation relevant to the experiment. The second section briefly describes some important properties of the Lithium-$7$ atom. Finally, in the third section, the cooling, trapping and state manipulation techniques used in the experiment are presented, including Zeeman slower (\ref{theory:zeeman}), MOT(\ref{theory:mot}), gray molasses (\ref{theory:gm}), magnetic trap (\ref{theory:mt}) and optical dipole trap (\ref{theory:odt}).

\section{Bose-Einstein Condensate in Harmonic Trap}\label{theory:bec}

Every real particles can be classified as one of the two families according to their spins, fermions which have half integer spins and bosons which have integer spins. According to quantum field theory\cite{spin-statistics1,spin-statistics2} the many-particles wave function of identical particles must be symmetric or anti-symmetric under particle exchange for bosons or fermions respectively. For bosonic particles, because of the symmetry of the wave function, the possibility for particles to be in the same state is greatly enhances. As a result, boson gas at ultra-low temperature forms a Bose-Einstein condensate (BEC), in which almost all of the particles are condensed to the lowest energy state. In the following, the relevant properties of BECs such as critical temperature and density distribution are described.\\
\\
Since the wave function of bosons is symmetric, multiple bosons can be in the same state. From this fact, the energy distribution can be calculated for bosons,
\eqar{
  f(\varepsilon)=&\frac{1}{\ue^{\beta(\varepsilon-\mu)} - 1}
}
Since the distribution has to be positive for all energy states, in particular the $\varepsilon=0$ ground state, we have $\mu\geqslant0$. For all the state except the ground state, this sets an finite upper limit on the number of atoms in each state for a fixed temperature. Therefore, if the number of atoms exceeds a certain value, all the extra atoms will go into the ground state. These atoms condensed in the ground state are called the Bose Einstein condensate.\\
\\
In order to calculate the atom number in the condensate as well as the critical temperature, we can estimate the maximum atom number in the excited states (thermal atoms) with an integral,
\eqar{
  N_{th}=&\int_0^\infty\frac{g(\varepsilon)}{\ue^{\beta\varepsilon} - 1}\ud\varepsilon
}
where $g(\varepsilon)$ is the energy density of states. In our experiment, we create the BEC in a harmonic optical dipole trap (see \ref{theory:odt}) for which the energy density is,
\eqar{
  g(\varepsilon)=&\frac{\varepsilon^2}{2\hbar^3 \omega^3}\\
  N_{th}=&\frac{1}{2\hbar^3 \omega^3}\int_0^\infty\frac{\varepsilon^2}{\ue^{\beta\varepsilon} - 1}\ud\varepsilon\\
  =&\frac{1}{2\hbar^3 \omega^3\beta^3}\int_0^\infty\frac{x^2}{\ue^{x} - 1}\ud x\\
  =&\frac{k_B^3T^3}{2\hbar^3 \omega^3}\zeta(3)\Gamma(3)
}
The critical temperature of the transition, determined by $N_{th}=N$,
\eqar{
  T_C=&\frac{\hbar\omega}{k_B}\sqrt[3]{\frac{2N}{\zeta(3)\Gamma(3)}}\\
  =&0.9405\frac{\hbar\omega\sqrt[3]{N}}{k_B}
}
Condensate fraction (for large $N$),
\eqar{
  \frac{N_0}{N}=&1-\frac{N_{th}}{N}\\
  =&1-\paren{\frac{T}{T}}^3
}

\section{Lithium-$7$ Atoms}

The atoms used in this experiment is the Bosonic isotope of Lithium, Lithium-$7$. In this section, I will present some properties of the Lithium-$7$ atoms that are important for the cooling and trapping techniques we use in the experiment.\\
\\
One of the most important properties for laser manipulation of cold atoms is the energy level structure. As all other alkali atoms, Lithium-$7$ has one unpaired valence electron with a $nS$ ground state and $nP$ excited states, where $n=2$ for Lithium-$7$. The fine structure split the excited states into $2^2P_{1/2}$ and $2^2P_{3/2}$ states corresponding to the $D1$ and $D2$ lines that are about $10GHz$ apart. On top of these, the hyperfine structure caused by the nucleus spin ($I=3/2$ for Lithium-$7$) further split each of these states into different levels with different total angular momentum $F$. the precise frequencies of these transitions measured in the experiment are shown in figure \ref{li:lines}\\
\begin{figure}
  \begin{center}
    \begin{tikzpicture}[scale=1.5]
      % S1/2
      % F1
      \draw[line width=1] (0, 0) -- +(5, 0) node[right] {$F=1$};
      % F2
      \draw[line width=1] (0, 1) -- +(5, 0) node[right] {$F=2$};
      \path (6, 0.5) node[right] {$2^2S_{1/2}$};

      % P1/2
      % F1
      \draw[line width=1] (0, 4) -- +(5, 0) node[right] {$F=1$};
      % F2
      \draw[line width=1] (0, 5) -- +(5, 0) node[right] {$F=2$};
      \path (6, 4.5) node[right] {$2^2P_{1/2}$};

      % P3/2
      % F3
      \draw[line width=1] (0, 6.5) -- +(5, 0) node[right] {$F=3$};
      % F2
      \draw[line width=1] (0, 7.5) -- +(5, 0) node[right] {$F=2$};
      % F1
      \draw[line width=1] (0, 8.1) -- +(5, 0) node[right] {$F=1$};
      % F0
      \draw[line width=1] (0, 8.5) -- +(5, 0) node[right] {$F=0$};
      \path (6, 7.5) node[right] {$2^2P_{3/2}$};

      % D1
      \draw[line width=1.5,->] (2, 5) -- (2, 1);
      \path (2, 3) node[left,align=right] {$D1$\\$446.799\ 862\ 994(12)$THz\\ ($670.977\ 058\ 925(18)$nm)};

      % D2
      \draw[line width=1.5,->] (3, 6.5) -- (3, 1);
      \path (3, 3) node[right,align=left] {$D2$\\$446.809\ 874\ 988(27)$THz\\ ($670.962\ 023\ 854(41)$nm)};

      % S1/2
      \draw[line width=1.5,->] (2.5, 1) -- (2.5, 0);
      \path (2.5, .5) node[right] {$803.504\ 086\ 6(10)$MHz};

      % P1/2
      \draw[line width=1.5,->] (1.5, 5) -- (1.5, 4);
      \path (1.5, 4.5) node[left] {$91.877(9)$MHz};

      % P3/2
      \draw[line width=1.5,->] (3.5, 7.5) -- (3.5, 6.5);
      \path (3.5, 7) node[right] {$9.462(27)$MHz};
      \draw[line width=1.5,->] (2.5, 8.1) -- (2.5, 7.5);
      \path (2.5, 7.8) node[right] {$5.813(27)$MHz};
      \draw[line width=1.5,->] (1.5, 8.5) -- (1.5, 8.1);
      \path (1.5, 8.3) node[right] {$2.759(27)$MHz};
    \end{tikzpicture}
  \end{center}
  \caption{Energy level structure of Lithium-$7$\cite{lithium-lines}}
  \label{li:lines}
\end{figure}\\
In an external magnetic field, each $F\neq0$ levels will be split because of the Zeeman effect. At low field, this splitting is described by the Land\'e $g$-factor $\Delta=g_F\mu_Bm_FB_z$ as shown in table \ref{li7:g-factors}. At higher magnetic field, different hyperfine levels starts to mix and the Zeeman effect is no-longer linear. Most importantly, for $2^2S_{1/2}$ ground states the $F=1$ states have a negative slope at high magnetic field and all the $F=2$ states except $m_F=-2$ has a positive slope. (Note that $F$ and $m_F$ are no-longer good quantum numbers at high magnetic field but there are states at high field that are adiabatically connected with these $F$ and $m_F$ eigenstates). The states with a positive slope of the Zeeman shift are magnetically trappable states since they can be trapped by a static magnetic field which can have a local minimum (but not a maximum \ref{theory:mt}) in the center. Only $|F=2, m_F=2\rangle$ and $|F=2, m_F=1\rangle$ states are both trappable at high field and low field.

\begin{table}
  \caption{Land\'e $g$-factors of Lithium-$7$}
  \label{li7:g-factors}
  \begin{center}
    \begin{tabular}{|c|c|c|}\hline
      Fine Structure & $F$ & $g$-factor \\\hline
      $2^2S_{1/2}$ & $2$ & $\dfrac 12$ \\\cline{2-3}
      & $1$ & $-\dfrac 12$ \\\hline
      $2^2P_{1/2}$ & $2$ & $\dfrac 16$ \\\cline{2-3}
      & $1$ & $-\dfrac 16$ \\\hline
      $2^2P_{3/2}$ & $1$, $2$, $3$ & $\dfrac 23$ \\\hline
    \end{tabular}
  \end{center}
\end{table}

\section{Cooling and Trapping Theory}

The atoms used in an alkali atom experiment often come from an atom beam coming out of a oven kept above the melting temperature of the metal. The oven used in this experiment operates at $485^\circ\text{C}$ producing an atom beam traveling at several hundred meters per second. In order to achieve low temperature and high density, several stages of slowing, trapping and cooling are implemented in this experiment which finally bring the atoms down to the Bose-Einstein condensate condition. In this section, I am going to talk about the theory behind these techniques we are using in our experiment.

\subsection{Zeeman Slower}\label{theory:zeeman}

The atoms comes out of the oven has an average velocity determined by the Maxwell-Boltzmann distribution,
\eqar{
  \bar v=&\sqrt{\frac{k_BT}{2\pi m}}
}
We can slow down the atoms using the recoil of photon scattering by shining resonance light to the beam. However, since the atomic transition resonance is very narrow ($\Gamma=2\pi\cdot5.9\text{MHz}$) compare to the Doppler shift ($\Delta\nu=\nu v/c\approx590\text{MHz}$), the atom will soon shift out of resonance once it slows down. There are several ways to solve this problem. One way is to modulate the laser frequency, either sweeping with time (frequency chirping) or making it broadband (white light). The Zeeman slower solves the problem by changing the resonance of the atom with the Zeeman effect. This is the most popular way for alkali atoms which usually have a cycling transition with a linear Zeeman effect in a large range of magnetic field.\\
\\
The Zeeman slower has a magnetic field aligned with the atom beam with a spatially varying amplitude. By shining circularly polarized light against the atom beam, the atoms in the beam are optically pumped to one of the two stretched states determined by the relative direction of the field and the light polarization. If the magnetic field is changing in a way such that the Zeeman shift of the atom follows the Doppler shift when atoms slow down, the atom beam will be always in resonance and continuously being slowed down, i.e.
\eqar{
  g_Fm_F\mu_BB=\frac{v}{c}\nu+\Delta
}
For constant acceleration, this implies,
\eqar{
  B=\frac{\sqrt{v_0^2-2ax}}{g_Fm_F\mu_Bc}\nu+B_0
}
The maximum deceleration of the Zeeman slower is limited by the maximum scattering rate, therefore the life time of the excited state,
\eqar{
  a_{max}=&\frac{h\Gamma}{2\lambda}
}
where $\Gamma$ is the linewidth of the transition.

\subsection{Magneto-Optical Trap (MOT)}\label{theory:mot}

As in most cold atom experiment, our experiment starts with loading Zeeman slowed atom into a magneto-optical trap (MOT). MOT is a technique that uses laser and magnetic field gradient to provides both molasses cooling and confinement. In order to understand how the MOT works, we will consider a one-dimensional model consisting of a spatially varying magnetic field $B=B_x=B'x$, two counter propagating, red-detuned, circularly polarized light with atoms sitting close to the zero of the magnetic field.\\
\\
The cooling and trapping effect in a MOT can be understood separately. Let us first consider an atom in the center of the MOT moving in the $+x$ direction. Since the light is red-detuned, the Doppler effect will shift the atom closer to the light in $-x$ direction and farther from the light in $+x$ direction. The atom will therefore scatters more photons from the $-x$ light and experience an average force proportional but in the opposite direction with its velocity. Such a force will damp the motion of the atom and therefore cool down the temperature of the cloud.\\
\\
In order to understand the source of the trapping force in a MOT, we can consider an atom with a $F=0$ ground state and $F=1$ excited states. As we can see in the schematic of this model (Figure \ref{mot:energy}), since the magnetic field changes its direction in the center, both laser beams are $\sigma^-$ light on the side it is coming from and therefore is shifted closer to resonance by the Zeeman effect. Therefore, atoms shifted from the center will scatter more photon from this direction and on average feel a net force from recoil pointing toward the center of the trap.\\
\begin{figure}
  \begin{center}
    \begin{tikzpicture}[scale=2]
      % axis
      \draw[line width=1.3,->] (-2.5, 0) -- (2.5, 0) node[below] {$x$};
      \draw[line width=1.3,->] (0, 0) -- (0, 3.5) node[right] {$\omega$};

      % Zeeman levels
      % F = 1
      \draw[line width=1] (-1.5, 2.5) -- (1.5, 2.5);
      \draw[line width=1] (-1.5, 3) -- (1.5, 2);
      \draw[line width=1] (-1.5, 2) -- (1.5, 3);

      % F = 0
      \draw[line width=1] (-1.5, .2) -- (1.5, .2);

      % Light freq
      \draw[line width=1,dashed] (-1.5, 1.8) -- (1.5, 1.8) node[below right] {Laser Frequency};

      % Zeeman labels
      \path (1.5, 3.5) node[right] {$|F, m_F\rangle$};
      \path (1.5, 3) node[right] {$|1, 1\rangle$};
      \path (1.5, 2.5) node[right] {$|1, 0\rangle$};
      \path (1.5, 2) node[right] {$|1, -1\rangle$};
      \path (1.5, .2) node[right] {$|0, 0\rangle$};

      \path (-1.5, 3.5) node[left] {$|F, m_F\rangle$};
      \path (-1.5, 3) node[left] {$|1, 1\rangle$};
      \path (-1.5, 2.5) node[left] {$|1, 0\rangle$};
      \path (-1.5, 2) node[left] {$|1, -1\rangle$};
      \path (-1.5, .2) node[left] {$|0, 0\rangle$};

      % Light polarization
      \draw[line width=1,snake arrow] (-1.5, 1.2) node[left] {$+x$ polarization} -- (1.5, 1.2);
      \draw[line width=2,->] (-1.2, 1.2) [partial ellipse=30:330:0.1 and 0.2];
      \path (-0.7, 1.2) node[above] {$\sigma^-$};
      \path (0.7, 1.2) node[above] {$\sigma^+$};

      \draw[line width=1,snake arrow] (1.5, 0.7) node[right] {$-x$ polarization} -- (-1.5, 0.7);
      \draw[line width=2,->] (1.2, 0.7) [partial ellipse=330:30:0.1 and 0.2];
      \path (0.7, 0.7) node[below] {$\sigma^-$};
      \path (-0.7, 0.7) node[below] {$\sigma^+$};

    \end{tikzpicture}
  \end{center}
  \caption{Schematic of a $1D$ MOT with a $F = 0$ to $F = 1$ transition.}
  \label{mot:energy}
\end{figure}\\
The performance of the MOT is usually limited by the random kick from scattered photons (Doppler limit on temperature) and light induced collisions (limits the density and atom number).

\subsection{Gray Molasses}\label{theory:gm}

Since the $2^2P_{3/2}$ excited state of Lithium-$7$ we use for molasses cooling in the MOT does not have resolved hyperfine structure (Figure \ref{li:lines}). We cannot preform conventional polarization gradient cooling using the $D2$ line. Instead, we implement a gray molasses on the D1 transition. This scheme was first studied and demonstrated by Christophe Salomon\cite{gm-theory} to achieve sub-Doppler temperature on some alkali isotopes that are not amenable for conventional polarization cooling ($^7Li$, $^6Li$ and $^{40}K$). In this section, I am going to present a semi-quantitative explanation of this cooling scheme. See the original paper\cite{gm-theory} for more detail discussion.\\
\\
The gray molasses cooling uses three pairs of counter-propagating beams that can cool the cloud in all three dimensions. Like what we did for MOT, we will consider a $1-D$ case for simplicity. The model we will use has a pair of counter-propagating beams, each has a high intensity pumper, with a large blue detuning $\delta$ from the $F=2$ ($|2\rangle$) to $F'=2$ ($|3\rangle$) $D1$ transition, and a low intensity repumper, blue detuned from the $F=1$ ($|1\rangle$) to $F'=2$ $D1$ transition by the same detuning $\delta$. Let the Rabi frequency for the pumper and the repumper be $\Omega_2(x)$ and $\Omega_1(x)$ respectively.\\
\\
Just as any other laser cooling method, the performance of the gray molasses is determined by the balance between cooling force and heating from photon scattering. Since the two beams (pumper and repumper) satisfies the Raman resonance condition, the system has a dark state which is the superposition of $|1\rangle$ and $|2\rangle$,
\eqar{
  |NC\rangle=&\frac{\Omega_2|1\rangle-\Omega_1|2\rangle}{\sqrt{\Omega_1^2+\Omega_2^2}}
}
which will be the state most of the atoms will be pumped into. (The state is dark for stationary atoms and has a non-zero but small scattering rate for moving atoms.) Since the scattering is suppressed by the present of a dark state, the gray molasses produce little heating.\\
\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      % axis
      \draw[line width=1.3,->] (-2, 0) -- (5, 0) node[below] {$x$};
      \draw[line width=1.3,->] (-2, 0) -- ++(0, 8) node[above] {$E$};

      % Dressed levels 1
      \draw[line width=2] plot[id=x,domain=1:4,samples=1000] function{3 + 0.2 * sin(5 * x)} node[right] {$|3',n-1\rangle$};
      \draw[line width=1] plot[id=x,domain=1:4,samples=1000] function{4.2 - 0.2 * sin(5 * x)};
      \draw[line width=1] plot[id=x,domain=1:4,samples=1000] function{4.215 - 0.215 * sin(5 * x)} node[right] {$|2',n\rangle$};
      \draw[line width=1] plot[id=x,domain=1:4,samples=1000] function{4.23 - 0.23 * sin(5 * x)};
      \draw[dashed] (1, 3.986) -- ++(3, 0);

      % Dressed levels 2
      \draw[line width=2] plot[id=x,domain=1:4,samples=1000] function{6.5 + 0.2 * sin(5 * x)} node[right] {$|3',n\rangle$};
      \draw[line width=1] plot[id=x,domain=1:4,samples=1000] function{7.7 - 0.2 * sin(5 * x)} node[right] {$|2',n+1\rangle$};
      \draw[line width=1] plot[id=x,domain=1:4,samples=1000] function{7.715 - 0.215 * sin(5 * x)};
      \draw[line width=1] plot[id=x,domain=1:4,samples=1000] function{7.73 - 0.23 * sin(5 * x)};
      \draw[dashed] (1, 7.486) -- ++(3, 0);

      % F1 states
      \draw (0.5, 0.5) -- ++(-0.5, 0) node[left] {$|1,n-1\rangle$};
      \draw (0.5, 3.5) -- ++(-0.5, 0) node[left] {$|1,n\rangle$};

      % Repumper
      \draw[red,line width=1,dashed] (pi / 2 - 0.25, 7.35) -- ++(0.5, 0);
      \draw[red,line width=1,snake arrow] (0.25, 3.5) -- (pi / 2, 7.35);
      \draw[red,line width=1,snake arrow] (pi / 2, 7.35) -- (pi / 2, 4);
      \draw[red,line width=2.6] plot[id=x,domain=pi / 2:(pi * 0.7 + 0.1),samples=1000] function{4.215 - 0.215 * sin(5 * x)};
      % Somehow math functions does not work correctly here.
      \draw[red,line width=1,snake arrow] (pi * 0.7 + 0.1, 4.215 + 0.215 * 0.87758256189037287) -- (0.25, 0.5);
    \end{tikzpicture}
  \end{center}
  \caption{Schematic of a $1D$ Gray Molasses.\cite{gm-theory}}
  \label{gm:dressed}
\end{figure}\\
The cooling effect of the gray molasses can be understood in the dressed atom picture for the atom and the pumper beam as shown in figure \ref{gm:dressed}, where the new energy eigenstates are,
\eqar{
  |2'\rangle=&\frac{\delta^2}{\delta^2+\Omega_2(x)^2}\paren{|2\rangle-\ui\Omega_2(x)|3\rangle}\\
  |3'\rangle=&\frac{\delta^2}{\delta^2+\Omega_2(x)^2}\paren{|3\rangle-\ui\Omega_2(x)|2\rangle}
}
Since the detuning $\delta$ is large, the mixing between $|2\rangle$ and $|3\rangle$ is small giving $|3'\rangle$ a short lifetime and $|2'\rangle$ a longer lifetime. The two counter-propagating pumper beams create a standing wave which causes periodic AC-Stark shift in the dressed $|2'\rangle$ and $|3'\rangle$ states. A repumper with the same or slightly smaller blue detuning compared to the pumper is closest to the resonance with the dressed states at the nodes of the standing wave where the mixing is minimized. Therefore, the repumper most likely pumps atoms into the $|2'\rangle$ state (through $|3'\rangle$ as shown in figure \ref{gm:dressed}) at the node of the standing wave and the atoms will most likely exit the state at the anti-node of the standing wave where the lifetime of $|2'\rangle$ is the shortest because of the maximum mixing. Since potential energy for atoms in $|2'\rangle$ is minimized at the nodes and maximized at the anti-nodes, this is a realization of the Sisyphus cooling scheme in which the atoms feel on average a damping force by climbing up the hill repeatedly in $|2'\rangle$. The height of the hill, therefore the cooling force, is proportional to the laser intensity. Combined with the low scattering rate due to the dark state, the gray molasses can be used to quickly cool the cloud after MOT to sub-Doppler temperature.

\subsection{Evaporation in a Static Magnetic Trap}\label{theory:mt}

The density and temperature of a laser cooled cloud is usually limited by the scattering of photons. Therefore, in order to reach quantum degeneracy, one usually needs to use a way that can cool the atoms without shining resonant light onto them. The method we use in our experiment is evaporative cooling in a magnetic trap and an optical dipole trap. In this section, I will describe the theory of static magnetic quadrupole trap and radio frequency evaporation and the theory of evaporation in an optical dipole trap will be discussed in the next section.\\
\\
The static magnetic trap, as the name suggested, is created using static magnetic field and the Zeeman effect of the atoms. Assuming a monotonic the Zeeman effect, the atoms will be trapped at an extremum of the field. Since the magnetic field in the vacuum satisfies the source-less Maxwell equation, we have,
\eqar{
  0=&\nabla\times\paren{\nabla\times B}\\
  =&\nabla\paren{\nabla\cdot B}-\nabla^2B\\
  =&-\nabla^2B\\
  \nabla^2\paren{B^2}=&2\nabla\paren{B\cdot\nabla B}\\
  =&2\paren{\nabla B}^2 + 2B\cdot\nabla^2B\\
  =&2\paren{\nabla B}^2\\
  \geqslant&0
}
Therefore, the only local extremum that is possible for a static magnetic field in the vacuum is a local minimum and only states with a positive Zeeman shift (low field seeking) can be trapped magnetically (known as Wing's theorem\cite{wing-theorem}).\\
\\
The static magnetic field with a local minimum in the center can be implemented in many different ways. The configuration used in this experiment is a quadrupole field created by a set of Helmholtz coils carrying opposite current, which provides the tighter confinement compare to other configurations. Closed to the center of the trap, the magnetic field can be expressed as,
\eqar{
  B_x=&B'x\\
  B_y=&B'y\\
  B_z=&-2B'z
}
Since the magnetic field is $0$ at the center of the trap where all the Zeeman levels are degenerate, spin-flip of atoms (Majorana loss) occurs in the center of a quadrupole trap. This problem is fixed in the experiment by plugging the hole with a repulsive blue detuned laser beam. This problem as well as the solution will be further discussed in the experimental chapter (\ref{exp:mt})\\
\begin{figure}
  \begin{center}
    \subfigure[The beginning of RF evaporation]{
      \begin{tikzpicture}[scale=0.5]
        \begin{scope}
          \clip plot[id=x,domain=-3.5:3.5,samples=1000] function{4 - (sqrt((abs(x) - 3)**2 + 1.0 / 9) - 3)} -- (3.5, 8) -- (-3.5, 8) -- cycle;
          \clip[shade, bottom color=red,top color=white] (3, 7 - 1.0 / 3) -- (-3, 7 - 1.0 / 3) -- (-3, 4) -- (3, 4) -- cycle;
        \end{scope}

        % axis
        \draw[line width=1,->] (-5, 0) -- (5, 0) node[below] {$x$};
        \draw[line width=1,->] (-0, 0) -- ++(0, 8) node[above] {$E$};

        % 2, 2
        \draw[dashed] plot[id=x,domain=-3:3,samples=1000] function{abs(x) + 4} node[right] {$|F=2, m_F=2\rangle$};

        % 1, 1
        \draw[dashed] plot[id=x,domain=-3:3,samples=1000] function{3 - abs(x)} node[above right] {$|F=1, m_F=1\rangle$};

        % 2, 2 bend
        \draw[color=blue] plot[id=x,domain=-3.5:3.5,samples=1000] function{4 - (sqrt((abs(x) - 3)**2 + 1.0 / 9) - 3)};

        % 1, 1 bend
        \draw[color=blue] plot[id=x,domain=-3.5:3.5,samples=1000] function{3 + (sqrt((abs(x) - 3)**2 + 1.0 / 9) - 3)};

        % RF
        \draw[color=red,<->,line width=2] (3, 0) -- (3, 7);
        \path (3, 3.5) node[right] {RF knife};
      \end{tikzpicture}
    }
    \subfigure[The ending of RF evaporation]{
      \begin{tikzpicture}[scale=0.5]
        \begin{scope}
          \definecolor{dark blue}{rgb}{0,0,0.8};
          \clip plot[id=x,domain=-3.5:3.5,samples=1000] function{4 - (sqrt((abs(x) - 1)**2 + 1.0 / 9) - 1)} -- (3.5, 8) -- (-3.5, 8) -- cycle;
          \clip[shade, bottom color=dark blue,top color=white] (1, 5 - 1.0 / 3) -- (-1, 5 - 1.0 / 3) -- (-1, 4) -- (1, 4) -- cycle;
        \end{scope}

        % axis
        \draw[line width=1,->] (-5, 0) -- (5, 0) node[below] {$x$};
        \draw[line width=1,->] (-0, 0) -- ++(0, 8) node[above] {$E$};

        % 2, 2
        \draw[dashed] plot[id=x,domain=-3:3,samples=1000] function{abs(x) + 4} node[right] {$|F=2, m_F=2\rangle$};

        % 1, 1
        \draw[dashed] plot[id=x,domain=-3:3,samples=1000] function{3 - abs(x)} node[above right] {$|F=1, m_F=1\rangle$};

        % 2, 2 bend
        \draw[color=blue] plot[id=x,domain=-3.5:3.5,samples=1000] function{4 - (sqrt((abs(x) - 1)**2 + 1.0 / 9) - 1)};

        % 1, 1 bend
        \draw[color=blue] plot[id=x,domain=-3.5:3.5,samples=1000] function{3 + (sqrt((abs(x) - 1)**2 + 1.0 / 9) - 1)};

        % RF
        \draw[color=blue,<->,line width=2] (1, 2) -- (1, 5);
        \path (1, 3.5) node[right] {RF knife};
      \end{tikzpicture}
    }
  \end{center}
  \caption{RF evaporation in magnetic trap. The temperature goes down and the phase space density goes up during evaporation.}
  \label{mt:evap}
\end{figure}\\
After trapping the atoms with the magnetic field, we use radio frequency induced spin-flip to evaporate the cloud. As shown in figure \ref{mt:evap}, the $|F=2, m_F=2\rangle$ has a positive slope of the Zeeman shift whereas the $|F=1, m_F=1\rangle$ state has a negative one. After all the atoms are trapped in the $|F=2, m_F=2\rangle$ state, a RF field that is blue detuned to the zero field resonance is applied which mixes the two states and turns the potential curve around at the resonance field. As a result, the depth of the trap is lowered and atoms which has higher energy can now escape from the trap. If the RF knife is placed much higher than the thermal energy $k_BT$, each leaving atoms will carry away energy a lot higher than the averaged value leaving a lower average energy, therefore lower temperature, for the atoms remaining in the trap. After the cloud gets colder, fewer and fewer atoms will have the energy to pass the RF knife, at which point the RF frequency can be sweep down to accelerate the evaporation. In order for the evaporative cooling to work efficiently, a high elastic collision rate (high rate for creating atoms with high energy) is also necessary.

\subsection{Evaporation in Optical Dipole Trap}\label{theory:odt}

Although RF evaporation in the magnetic trap is very convenient and efficient, it does have some limitations,
\begin{enumerate}
\item
  The atoms have to be in a specific hyperfine state.
\item
  Cannot use a Feshbach resonance to tune the interaction with magnetic field.
\end{enumerate}
Moreover, for Lithium-$7$, the trappable $|F=2, m_F=2\rangle$ state has a negative scattering length ($-27a_0$) making it impossible to create a stable Bose-Einstein condensate with a large atom number. In order to solve the problem, we switch to an optical dipole trap and do the final evaporation there.\\
\\
The optical dipole trap (ODT) is created using the negative AC Stark shift of a red detuned laser beam. Since the AC Stark effect is proportional to $\delta^{-1}$ (where $\delta$ is the detuning) and the scattering (heating) is proportional to $\delta^{-2}$, high power beams with huge detuning ($\approx100THz$) are often used in order to introduce the least amount of heating for the same trapping potential.\\
\\
Evaporation in an ODT is done simply by lowering the potential (beam power). Since a positive and big scattering length is necessary for efficient evaporation to BEC, we spin-flip the atoms into $|F=1, m_F=1\rangle$ and evaporate close to a Feshbach resonance with an appropriate scattering length.
